Scalable backpropagation for Gaussian Processes using celerite

### 2018-01-30 paper/ms.tex
title
submitted version

\title{%
    Scalable gradients of Gaussian Process models with celerite
}

@@

GPs \citep{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some examples include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017, Angus:2018}, data-driven
models of light curves or stellar spectra \citep{Wang:2012, Luger:2016,
Czekala:2017}, and the cosmic microwave background
\citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is generally the computational bottleneck.
The details of these models are omitted here (see \citealt{Rasmussen:2006} and
\citealt{Foreman-Mackey:2017} for details), but the key point is that, for a
dataset with $N$ data points, every evaluation of a GP model requires
computation of the inverse\footnote{Or, more specifically, a linear operator
that can left multiply a vector or a matrix by the inverse of this matrix.}
and log-determinant of the $N \times N$ covariance matrix $K_\bvec{\alpha}$.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Ambikasaran:2015, Foreman-Mackey:2017}.

@@

\noindent In this algorithm, the \texttt{zeros}$(J,\,K)$ function creates a $J
\times K$ matrix of zeros, the \texttt{diag} function creates a diagonal
matrix from a vector, and $\bvec{x}_n$ indicates a row vector made from the
$n$-th row of the matrix $X$.
The computational cost of this algorithm scales as $\mathcal{O}(N\,J^2)$.
Using this factorization, the log-determinant of $K$ is
\begin{eqnarray}
\log \det K &=& \sum_{n=1}^N \log d_{n} \quad.
\end{eqnarray}
Similarly, \citet{Foreman-Mackey:2017} derived a $\mathcal{O}(N\,J)$ algorithm
to apply the inverse of $K$ as follows (Equations~47 and~48 in
\citealt{Foreman-Mackey:2017}):

@@

\section{Discussion}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figure.pdf}
\caption{%
    The empirical computational scaling of the algorithms presented in this
    note.
    \emph{(top row)}: The cost of computing \eq{loglike} as a
    function of the number of data points ($N$; left) and the model complexity
    ($J$; right).
    \emph{(bottom row)}: The cost of computing the gradient of \eq{loglike}
    with respect to the vector $\bvec{a}$ and the matrices $U$, $V$, and $P$.
    In the left panels, each line corresponds to a different value of $J$ as
    indicated in the legend (with $J$ increasing from bottom to top).
    Similarly, in the right panels, the lines correspond to different values
    of $N$ increasing from bottom to top.
    In both cases, the theoretical scaling is $\mathcal{O}(N,J^2)$.
\figurelabel{figure}}
\end{center}
\end{figure}

>>>>

\title{%
    Scalable backpropagation for Gaussian Processes using celerite
}

@@

GPs \citep{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some examples include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017, Angus:2018}, data-driven
models of light curves or stellar spectra \citep{Wang:2012, Luger:2016,
Czekala:2017}, and the cosmic microwave background
\citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is generally the computational bottleneck.
The details of these models are omitted here (see \citealt{Rasmussen:2006} and
\citealt{Foreman-Mackey:2017} for details), but the key point is that, for a
dataset with $N$ data points, every evaluation of a GP model requires
computing the log-determinant and multiplying a vector by the inverse of
%\footnote{Or, more specifically, a linear operator
%that can left multiply a vector or a matrix by the inverse of this matrix.}
the $N \times N$ covariance matrix $K_\bvec{\alpha}$.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Ambikasaran:2015, Foreman-Mackey:2017}.

@@

\noindent In this algorithm, the \texttt{zeros}$(J,\,K)$ function creates a $J
\times K$ matrix of zeros, the \texttt{diag} function creates a diagonal
matrix from a vector, and $\bvec{x}_n$ indicates a row vector made from the
$n$-th row of the matrix $X$.
The computational cost of this algorithm scales as $\mathcal{O}(N\,J^2)$.
Using this factorization, the log-determinant of $K$ is
\begin{eqnarray}
\log \det K &=& \sum_{n=1}^N \log d_{n} \quad.
\end{eqnarray}
Similarly, \citet{Foreman-Mackey:2017} derived a $\mathcal{O}(N\,J)$ algorithm
to apply the inverse of $K$ (i.e.\ compute $Z = K^{-1}\,Y$) as follows
(Equations~47 and~48 in \citealt{Foreman-Mackey:2017}):

@@

\newpage
\section{Discussion}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figure.pdf}
\caption{%
    The empirical computational scaling of the algorithms presented in this
    note.
    \emph{(top row)}: The cost of computing \eq{loglike} as a
    function of the number of data points ($N$; left) and the model complexity
    ($J$; right).
    \emph{(bottom row)}: The cost of computing the gradient of \eq{loglike}
    with respect to the vector $\bvec{a}$ and the matrices $U$, $V$, and $P$.
    In the left panels, each line corresponds to a different value of $J$ as
    indicated in the legend (with $J$ increasing from bottom to top).
    Similarly, in the right panels, the lines correspond to different values
    of $N$ increasing from bottom to top.
    In both cases, the theoretical scaling is $\mathcal{O}(N,J^2)$.
\figurelabel{figure}}
\end{center}
\end{figure}
